Using News' Title and Body Text Features to Detect Fake News: a Viability Study

### 2019-09-04 FakeNewsAnalysis/paperLatex/paperFelipeNevesDataScience.tex
Added plot script

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{multirow}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\begin{document}

@@

\begin{table}[htbp]
\caption{Features used in Horne's study}
\begin{center}
\begin{tabular}{ |l|l|l| }
\hline
Section & \multicolumn{2}{ c| }{Feature Sets} \\
\hline
\multirow{4}{*}{Title} & LB & Lucas Radebe \\
    & DC & Michael Duburry \\
    & DC & Dominic Matteo \\
    & RB & Didier Domi \\ \hline
\multirow{3}{*}{Body} & MC & David Batty \\
    & MC & Eirik Bakke \\
    & MC & Jody Morris \\ \hline
\end{tabular}
\label{table:horne_features}
\end{center}
\end{table}

In this paper he also implemented a SVM classifier to test the result of these features on his own
proposed dataset. Considering that he only used 4 features, which is a considerably low number for
a text classifier, he achieved a very impressive result with over 78\% accuracy in some tests , as
shown in Table \ref{table:horne_results}.

\begin{table}[htbp]
\caption{Results achieved in Horne's study}
\begin{center}
\begin{tabular}{ |l|l|l| }
\hline
Feature Set & Baseline & Accuracy \\
\hline
Title & 50\% & 71\% \\
Body & 50\% & 78\% \\
\hline
\end{tabular}
\label{table:horne_results}
\end{center}
\end{table}

@@

@@

\section{Discussion}

>>>>

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{multirow}
\usepackage{hyperref}

\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\begin{document}

@@

\begin{table}[htbp]
\caption{Features used in Horne's study}
\begin{center}
\begin{tabular}{ |l|l|l| }
\hline
Section & \multicolumn{2}{ c| }{Feature Sets} \\
\hline
\multirow{4}{*}{Title} & per\_stop & Percentage of stop words\\
    & NN & Number of nouns in the text \\
    & avg\_wlen & Average word length \\
    & FK & Flesch-Kincaid readability index \\ \hline
\multirow{4}{*}{Body} & NN & Number of nouns in the text \\
    & TTR & Token to type ration (lexical richness)\\
    & WC & Number of words in the text \\
    & quote & Number of quotes in the text \\ \hline
\end{tabular}
\label{table:horne_features}
\end{center}
\end{table}

In this paper he also implemented a \textit{SVM} classifier to test the result of these features on his own
proposed dataset. Considering that he only used 4 features, which is a considerably low number for
a text classifier, he achieved a very impressive result with over 78\% accuracy in some tests , as
shown in Table \ref{table:horne_results}.

\begin{table}[htbp]
\caption{Results achieved in Horne's study}
\begin{center}
\begin{tabular}{ |l|l|l| }
\hline
Feature Set & Baseline & Accuracy \\
\hline
Title & 50\% & 78\% \\
\hline
Body & 50\% & 71\% \\
\hline
\end{tabular}
\label{table:horne_results}
\end{center}
\end{table}

@@

For the development of this project we utilized the \textit{Python} language, with the aid of some external
libraries as described in the Table \ref{table:python_libs}. \textit{Python} was chosen because it is widely
considered the most accessible language for Data Science and it allows more people to be able to reproduce
the results presented here \cite{cielen_meysman_mohamed_2016}\cite{richard_2018}. The full source
code will also be made available at \url{https://github.com/Feupos/DataScience} for this purpose.
This is a very important topic and will be further discussed in Section \ref{section:discussion}.

\begin{table}[htbp]
\caption{Python Libraries Utilized}
\begin{center}
\begin{tabular}{ |c|c| }
\hline
Library & Purpose \\ \hline
Pandas & File processing and data structures \\ \hline
scikit-learn & Machine learning algorithms \\ \hline
NLTK & Natural language processing tools and dictionaries \\ \hline
SentiStrength & Text polarity evaluation tool \\ \hline
LexicalRichness & Lexical richness evaluation tool (TTR) \\ \hline
Seaborn & Data visualization \\ \hline
\end{tabular}
\label{table:python_libs}
\end{center}
\end{table}

With these tools we developed two scripts, one for extracting the features and another for testing
machine learning algorithms for the classification of Fake News. Two types of classifiers were used:
a linear kernel SVM (Space Vector Machine) classifier and a Random Forest classifier.

We used \textit{scikit-learn}'s \textit{k-fold} cross validation for the performance evaluation of
the classifiers.

Besides the features already mentioned in \ref{table:horne_features} an additional set of features
was used for the evaluating the performance of the algorithm, these features are listed on Table
\ref{table:expanded_features} and will be referred as 'expanded feature set' whereas the features from
Table \ref{table:horne_features} will be referred as 'original feature set'. In the same way the dataset
from Horne's work \cite{horne_2017} will be referred as 'original dataset' and the dataset from Szpakowski
\cite{szpakowski_2019} wil be referred as 'expanded dataset'.

\begin{table}[htbp]
\caption{Expanded Feature Set}
\begin{center}
\begin{tabular}{ |l|l|l| }
\hline
\multicolumn{2}{ |c| }{Features} \\
\hline
per\_stop & Percentage of stop words\\
NN & Number of nouns in the text \\
avg\_wlen & Average word length \\
FK & Flesch-Kincaid readability index \\
TTR & Token to type ration (lexical richness)\\
WC & Number of words in the text \\
quote & Number of quotes in the text \\
NNP & Number of proper nouns in the text \\
str\_neg & Average negative polarity of the text \\
str\_pos & Average positive polarity of the text \\
JJR & Number of comparative adjectives in the text \\
JJS & Number of superlative adjectives in the text \\
RBR & Number of comparative adverbs in the text \\
RBS & Number of superlative adverbs in the text \\
\hline
\end{tabular}
\label{table:expanded_features}
\end{center}
\end{table}

@@

During our tests we used multiple combinations of datasets and feature sets, to be able to highlight the
difference in performance between them. We have also divided our results in 3 categories, one using only
features extracted from the tiles, another with only features extracted from the bodies, and a third one
combining all the extracted features from both titles and bodies.

\subsection{Original Feature Set}

In this test we used both the original feature set and the SVM classifier, to be able to reproduce the results
found in Horne's work \cite{horne_2017}, however we were not able to achieve that, instead the results were
'inverted', meaning that the accuracy using the Title features was achieved using the Body features and vice versa.
This can be observed comparing our results from Table \ref{table:od_of} and the original result from Table \ref{table:horne_results}.

We can also notice how the performance was significantly worse when using the expanded dataset, with the results
on Table \ref{table:ed_of}.

\begin{table}[htbp]
\caption{Original Dataset and Original Features Performance}
\begin{center}
\begin{tabular}{ |l|l|l| }
\hline
Feature Set & Baseline & Accuracy \\
\hline
Title & 50\% & 72.67\% \\
\hline
Body & 50\% & 78.67\% \\
\hline
Combined & 50\% & 80\% \\
\hline
\end{tabular}
\label{table:od_of}
\end{center}
\end{table}

\begin{table}[htbp]
\caption{Expanded Dataset and Original Features Performance}
\begin{center}
\begin{tabular}{ |l|l|l| }
\hline
Feature Set & Baseline & Accuracy \\
\hline
Title & 50\% & 58.13\% \\
\hline
Body & 50\% & 64.27\% \\
\hline
Combined & 50\% & 60.27\% \\
\hline
\end{tabular}
\label{table:ed_of}
\end{center}
\end{table}

\subsection{Expanded Feature Set}

We also tried increasing the number of features in order to try to improve the performance,
specially in the expanded dataset that had a very poor performance with the original features.
In these tests we also switched the classifier for the Random Forest classifier due to the
SVM being to slow to calculate with too many features and entries. Previous tests had already
suggested that the classifier had little impact in the overall performance.
The results are shown in the Table \ref{table:od_ef} and Table \ref{table:od_of}.

\begin{table}[htbp]
\caption{Original Dataset and Expanded Features Performance}
\begin{center}
\begin{tabular}{ |l|l|l| }
\hline
Feature Set & Baseline & Accuracy \\
\hline
Title & 50\% & 71.33\% \\
\hline
Body & 50\% & 82\% \\
\hline
Combined & 50\% & 82\% \\
\hline
\end{tabular}
\label{table:od_ef}
\end{center}
\end{table}

\begin{table}[htbp]
\caption{Expanded Dataset and Expanded Features Performance}
\begin{center}
\begin{tabular}{ |l|l|l| }
\hline
Feature Set & Baseline & Accuracy \\
\hline
Title & 50\% & 61.93\% \\
\hline
Body & 50\% & 73.93\% \\
\hline
Combined & 50\% & 71.8\% \\
\hline
\end{tabular}
\label{table:ed_ef}
\end{center}
\end{table}

\subsection{Compared Results}

TODO: plot new graph

\section{Discussion}
\label{section:discussion}

\section{Discussion}
\label{section:discussion}

\section{Conclusion}
